\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2023-03-09 sections/chameleon.tex
Minor changes

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both forms, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. There are cases where a programmer wants the debugger to simply ``show the answer", and others to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.

% \begin{table}
%     \centering
\begin{scriptsize}
\begin{center}
    \begin{tabular}{ l l  }
     \textit{Mode} & \textit{Features} \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \end{center}
\end{scriptsize}
%     \caption{\chameleon{} modes and features}
%     \label {tab:chameleon-features}
% \end{table}

>>>>

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both forms, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. There are cases where a programmer wants the debugger to simply ``show the answer", and others to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are summarized in table~\ref{tab:chameleon-features}.%shown in table~\ref{tab:chameleon-features}.

\begin{table}
    \centering
\begin{scriptsize}
\begin{center}
    \begin{tabular}{ l l  }
     \textit{Mode} & \textit{Features} \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \end{center}
\end{scriptsize}
    \caption{\chameleon{} modes and features}
    \label {tab:chameleon-features}
\end{table}

### 2023-03-09 sections/discussion.tex
Minor changes

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output. In this section, we will discuss a few interpretations of the results.

@@

\subsection{Externalize Intermediate Typing Information}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although  these two statements are equivalent in theory, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

>>>>

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examines   the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using the \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output. In this section, we will discuss a few interpretations of the results.

@@

\subsection{Externalize Intermediate Typing Information}
We speculate another factor of the effectiveness of \chameleon{} interactive debugging tools is that they help programmers effectively chunk intermediate information. With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although  these two statements are equivalent in theory, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

### 2023-03-09 sections/evaluation.tex
Minor changes

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started. 
All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. They was then asked to correct the code with the help of the given tool.

@@

\textbf{RQ2.1}. The most striking feature of the data is that users tend to vary wildly in their use of the tool. Some users used the features extensively, while others completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups:

\begin{small}
\noindent\begin{tabularx}{\linewidth}{ 
  | >{\hsize=.3\hsize}X 
  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }

    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse. Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I. = [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I. = [11.96, 26.67]). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by \chameleon{}. This reduced the apparent benefit of \chameleon{} in these tasks. On the other hand, task 9 is the lengthiest task of all. It also involves deeply nested type definations that are harder to follow in mind.

@@

Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs. 47\% staying in the same mode. Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

>>>>

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started. 
All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. They were then asked to correct the code with the help of the given tool.

@@

\textbf{RQ2.1}. The most striking feature of the data is that users tend to vary wildly in their use of the tool. Some users used the features extensively, while others completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups in table~\ref{tab:interaction-level}.
\begin{table}
    \centering
\begin{scriptsize}
\begin{small}
\noindent\begin{tabularx}{\linewidth}{ 
  | >{\hsize=.3\hsize}X 
  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }
    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}
\end{scriptsize}
    \caption{Levels of programmer interaction and their description}
    \label {tab:interaction-level}
\end{table}

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse. Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I. = [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I. = [11.96, 26.67]). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by \chameleon{}. This reduced the apparent benefit of \chameleon{} in these tasks. On the other hand, task 9 is the lengthiest task of all. It also involves deeply nested type definitions that are harder to follow in mind.

@@

Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs. 47\% staying in the same mode. Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

### 2023-03-09 sections/related-work.tex
Minor changes

\subsection{Producing high-quality error explanation}
One weakness of compiler error messages, in general, is that they fail to explain the error in human language. As put in~\cite{barik_developers_2017}, ``Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message). Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors. Decafe~\cite{becker_effective_2016} is a tool that can rephrase Java compiler error messages into an enhanced version. In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices. Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory. In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.  These works show the significance of improving the language in the compiler error messages. Most principles and suggestions are followed in \chameleon{} in constructing error statements. However, these earlier studies were not targeting type errors alone but general compiler errors (some even include run time errors). The nuances of type errors, such as alternative typing, were not considered. Moreover, these explanation systems were designed specifically for novice users.

@@

Modern programming tools can offer alternative methods of code authoring, display real-time feedback and reveal complex programming contexts through visualizations. Many tools aim to improve the debugging experience using such capabilities. We list two. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (called ``strategies'' by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like ``why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. These debugging tools are important motivations for developing \chameleon{}. However, they focus on different aspects of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.

>>>>

\subsection{Producing high-quality error explanation}
One weakness of compiler error messages, in general, is that they fail to explain the error in human language. As put in~\cite{barik_developers_2017}, ``Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message). Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors. Decaf~\cite{becker_effective_2016} is a tool that can rephrase Java compiler error messages into an enhanced version. In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices. Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory. In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.  These works show the significance of improving the language in the compiler error messages. Most principles and suggestions are followed in \chameleon{} in constructing error statements. However, these earlier studies were not targeting type errors alone but general compiler errors (some even include runtime errors). The nuances of type errors, such as alternative typing, were not considered. Moreover, these explanation systems were designed specifically for novice users.

@@

Modern programming tools can offer alternative methods of code authoring, display real-time feedback and reveal complex programming contexts through visualizations. Many tools aim to improve the debugging experience using such capabilities. We list two. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (called ``strategies'' by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like ``why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. These debugging tools are important motivations for developing \chameleon{}. However, they focus on different aspects of the debugging process. Java Whyline mainly tackles the problem of unintended runtime behavior, while Hazel Tutor specializes in development assistance supported by type holes.
